Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-19 projectReport/chapter5.tex
Rereading the project. On 5.4

In this chapter we turn to the implementation of the function $G_T$ by discussing a method to learn the function $\Gamma$. We apply our methodology to some chaotic attractors. Most notably, we consider the reconstruction of the chaotic attractor of the double pendulum -- an  attractor that has not been successfully reconstructed because of its intricate structure. In fact the previous data-driven approaches also have not been successful in building a model from the double pendulum.

@@

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\ref{eqns_from_data}. To reach this point, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value and lies within the range $[-1,1]$ and required zero-padding (see paragraph below for further discussion), we continue by driving the system with external input through the function $g$.

From~\eqref{eqn_driving} we recall the equation of the RNN: $g(u,x) = (1-a) x + a \tanh(Au + \alpha B x)$. Recall that $g$ is SI-invertible when $A$ is invertible and has same dimension as $B$. We refer to $A$ as the input matrix and $B$ the reservoir matrix.

We make use of a Recurrent Neural Network (RNN) in our implementation to project the data on to a higher dimensional space, but we do not train the RNN in the conventional manner. \ednote{B: Should we say what the conventional manner is?} A few words about the RNNs: RNNs are widely used in the literature for temporal processing of data as a type of deep learning technique. They take the form of equations considered to be discrete-time version of continuous dynamical systems modelling a neural network of neurons~\cite{jaeger2001echo}.

We implement a feedforward network, (a network with no memory, unlike the RNN) required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library 
% \ednote{B: Is this something you want me to write in? M: We do not implement the RNN, but the FFN that learns Gamma through Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data. RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

If the RNN we will be using in the next step has $N$ neurons, then $A$ and $B$ are each $N\times{N}$ matrices. Inputs $u_n$ of dimension $K < N$ are padded by zeroes (as remarked earlier) to embed the input in a space of dimension $N$. As we do not need the entire left-infinite history of input, an arbitrary initialisation of the system with initial values followed by the driving of the system with the function $g$ means that the generated values will approach the actual solution elements in a uniform manner thanks to the UAP that we had discussed in the preceding chapter.

Thirdly, we prepare the network by setting up a network with $N$ neurons (the same value $N$ as in the dimensions of the matrices $A,B$). Learning of $\Gamma$ is done using the Adam Optimiser where the means square error (MSE) loss function is optimised. 
The system is trained several times (in our case 4) with incrementally smaller learning rates on each run (0.001 and thereafter divided by 10 at each run). 
Additional parameters such as the number of network-layers, neurons per layer and training-length are presented case-by-case below.

Once $\Gamma$ is learnt then reach the prediction step, where we forecast $u_n$ from $(x_{n-1},x_{n})$. Note that  due to causal embedding $(x_{n-1},x_{n})$ is in one-to-one correspondence with the left-infinite orbit $(\ldots,u_{n-1},u_{n-1})$. Once $u_n$ is computed using $\Gamma$, we feed that to the driven system $g$ along with  $x_n$ to obtain $x_{n+1}$. Now we apply $\Gamma$ on $(x_n,x_{n+1})$ to forecast $u_{n+1}$ and the whole process repeats. Note that this corresponds to implementing the two model equations for learning from data~\eqref{eqns_from_data}.
% We then reach the prediction step. Making use of $g$'s SI-invertibility and $g$ as defined in~\eqref{eqn_driving}, we have the following expression for $u_n$ replicated from~\eqref{eqn_SI_RNN}.
% \begin{equation*}
%   u_n := A^{-1}\bigg(\overline{\tanh}^{^{-1}}\frac{1}{a}\Big(x_{n+1}-(1-a)x_n\Big) \bigg) - \alpha B x_n
% \end{equation*}

% The two model equations for learning from data~\eqref{eqns_from_data} allows us to determine what the state $x_{n+1}$ at time $n+1$ will be. 
% We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

We also consider a more general problem of learning an attractor of dynamical system when only observations of an orbit are given. Explicitly, if $(W,T)$ is a dynamical system with dynamics generated by $w_{n+1}=Tw_n$, and if $\theta:W \to \mathbb{R}$ is an observable, then the task is to learn a dynamical system that is topologically conjugate to $(W,T)$ and predict $\theta(w_{m+1}),\theta(w_{m+2}),\ldots$ using the data $\theta(w_{0}),\theta(w_{1}),\ldots,\theta(w_{m})$.  So how do we do it?
Suppose the input generated from the delay-coordinate map $\Phi_{\theta,2d}(\theta(w_{n})) := (\theta(w_{n-2d}),\ldots,\theta(w_{n-1}),\theta(w_{n}))$ is such that Takens delay embedding theorem (see Theorem~\ref{Thm_Takens}) holds, in which case, there exists a homeomorphism  $F_\theta: \Phi_{\theta,2d}(\theta(w_{n})) \mapsto \Phi_{\theta,2d}(\theta(w_{n+1}))$. Then if we feed  the input values $u_n := \Phi_{\theta,2d}(\theta(w_{n}))$ to a driven system $g$ that is SI-invertible and has USP, the induced system $(Y_F,G_F)$ would be topologically conjugate to the inverse-limit space of
$(\Phi_{\theta,2d}(W), F_\theta)$ due to Theorem~\ref{Thm_CET}, and hence one could forecast $u_n,u_{n+1},\ldots$. In summary, one could forecast  the values $\theta(w_n), \theta(w_{n+1}),\ldots$ by feeding these observations as delay coordinates since the input is again an orbit from a dynamical system. 
The advantage of feeding delay-coordinates to a driven system  over learning a map that describes the evolution of the delay coordinates is that the embedding is stable in the sense of global dissipativity that we have described in the previous chapter, i.e., there is an attractor $A$ of a map that implements $G_T$ on an expansion $Y_F^+$ providing some robustness to noise.

Now discuss on another practical issue with forecasting chaotic dynamical systems.  
We note that the twin equations~\ref{eqns_from_data} generate a model from data. For a chaotic system, a small amount of noise due either to the input-noise or computational error would cause the solutions from even an accurate model with the same initial conditions to diverge. However, statistically, a chaotic system would behave well in the sense that the visiting frequency or the average time spent by an orbit in any region of the phase space of the dynamical system is identical for almost all initial conditions (almost all is specified as a measure), a result in ergodic theory (\cite{wikiErgodicTheory}) a topic beyond the scope of this thesis.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modeling the data we plot the densities of the trajectories of the actual and predicted solutions in~\ref{noisydp_success_density, fig:dp_success_density}.

In our implementation, the input and reservoir  matrices $A$ and $B$ in the RNN are sparse matrices. This sparsity is not essential but seems to perform slightly empirically better than choosing the matrices arbitrarily.  We also discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opt to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed.  We do not have any theoretical justifications for this.

%By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input.  In so doing we then also guarantee another form stability. \ednote{B: Not complete enough. Edit further and cite M: Please feel free to cite a wikipedia page}

% Lastly, we ensure the matrix $B$ has unit spectral radius. This is necessary due to the Causal Embedding Theorem~\ref{Thm_CET} which can then guarantee the existence of semi-conjugacy between the systems $(Y_T, G_T)$ and $(\widehat{U}_T, \widehat{T})$. In fact, the theorem actually requires that $\alpha{B}$ have spectral radius $<1$, so we impose the restriction $\alpha<1$ and consider only the matrix $B$'s spectral radius from thereon.
Lastly, we ensure the matrix $B$ has unit spectral radius so that the variable $\alpha$ in the RNN equation \eqref{eqn_driving} can be used to know if the USP is satisfied. It is a  feature that a matrix with a spectral radius $<1$ generally has the USP \cite{yildiz2012re}.

We remark on one more aspect that helps produce slightly better empirical results.  To learn $\Gamma: (x_{n-1},x_{n}) \mapsto u_n$ through our feedforward network, we could feed the principal components. 
Principal components are used only for a more efficient state representation to possibly reduce learning errors; it is not for a lossy approximation since we use all principal components.
More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. If $X_{1:N}=U\Sigma P^T$ here represents the singular value decomposition ($U$ and $V$ are unitary matrices and $\Sigma$ is a diagonal matrix) of $X_{1:N}$ and we refer to \cite{wikiSVdecomp}. We then denote the principal component matrix $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P$.
These principal components are then used to train a feedforward neural network. Denote the row vectors of $Z_{1:N}$ by 
$z_i^T, i=1,2,\ldots,N$ and the neural network by $NN$, the network is subsequently trained to learn an approximation of the map
$NN: (z_{n-1},z_n) \mapsto  u_n.
$
The learnt approximation of $NN$ can be used to approximate $\Gamma$ since 
\begin{equation}\label{Seq_RNN}
  NN\left( \begin{bmatrix} 
    P^Tx_{n-1} \\
    P^Tx_n
    \end{bmatrix}
    \right) = NN \circ 
    \begin{bmatrix}
    P^T & 0 \\
    0 & P^T 
    \end{bmatrix}\begin{bmatrix}
    x_{n-1}\\
    x_n
    \end{bmatrix} = u_n = \Gamma(x_{n-1},x_n).
\end{equation}

\section{Experimental Results}
In this section we showcase numerical results pertaining to 3 attractors that were simulated using the theory and methodology described. We first consider a simple physical system that exhibits complicated dynamics.
\subsection{Double Pendulum}

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See below the Figure~\ref{fig:dp_setup})
This system shows chaos strikingly. Consider again Figure~\ref{fig:dp_sdic}.
The system has four variables: angle and angular velocity of each pendulum rod. An ideal system is assumed where the rods are massless and no friction in the pivots are considered.
\begin{figure}[ht]
  \includegraphics[scale=0.35]{Graphs/_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP. Image taken from~\cite{DPSetup}}
\label{fig:dp_setup}
\end{figure}

The scalar equations of motion are derived using Newtonian physics and are reproduced from~\cite{DPFormulas} in Equations~\ref{eqns_dp}. Each pendulum component has an associated mass of the head, $m$, and length of the rod, $L_i$. Acceleration due to gravity is denoted by $g$ and initial velocities by $\dot{\theta}$
\begin{eqnarray}\label{eqns_dp}
  \ddot{\theta_{1}}  = \frac{-g(2m_1+m_2)\sin{\theta_1} - m_2g\sin(\theta_1-2\theta_2) - 2\sin(\theta_1-\theta_2)m_2({\dot{\theta_{2}}}^{2}L_2 + {\theta_{1}'}L_1\cos(\theta_1-\theta_2))} {L_1(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
  \\
  \ddot{\theta_{2}} = \frac{2\sin(\theta_1-\theta_2)(\theta_{1}'^{2}L_1(m_1+m_2) + g(m_1+m_2)\cos\theta_1 + \dot{\theta_{2}}^{2}L_2m_2\cos(\theta_1-\theta_2))}{L_2(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
\end{eqnarray}

% The second-order system is converted to a first-order ODE system and solved using the Runge-Kutta numerical method.  
Here we chose the values $m_1=2.0$, $m_2=2.0$, $L_1=1.5$ and $m_2=1.5$ with initial velocities $v_1=1.5, v_2=-2$ for the respective pendulum heads.
With a change of variables the second-order system can be 
written as a system of first order equations with four variables, where two variables devoted to each pendulum denote the $x$ and $y$ coordinates of the mass (the lengthy equations are not shown here). The ODE system is then solved using the fourth-order Runge-Kutta numerical method.

@@

We need the assumption the system is not damped so that the map $T$ resulting from the discretization of the system  is surjective.  If we assume damping surjectivity is lost on any phase space with more than one point.  
An example of this would be the case of a DP exhibiting energy-loss. In such a case, the attractor is a single point in the phase space when the double pendulum comes to rest. 
For such a system, it may be noted that if we use data outside the attractor, learning is not reliable. To see this, consider the plots ~\ref{fig:damped_pendulum} and ~\ref{fig:dp_notsurjective}.

@@

The system was driven for 15500 steps, of which the first 500 were immediately discarded to simulate the network `forgetting'. The driving function $g$ had input and reservoir matrices were implemented as sparse matrices filled to 20\%. Parameter-values of $\alpha=0.99$ and $a=0.5$ were used.
The remaining 15000 datapoints were used to train a network with 128 neurons. 
The network was initialised to contain 16 hidden layers each with a layer dimension of 64, activated with the $\tanh$ activation function.
Training was accomplished using 256 epochs with batch sizes 128 and a delay of 1. See Fig.~\ref{fig:dp_success_traj} for results pertaining to the trajectory taken by the DP's components and then Fig.~\ref{fig:dp_success_density} to witness associated densities of predicted versus actual values.

@@

We now consider an experiment where noise is added to the system and find it to be incredibly robust. We employ the same parameters as in the preceding experiment and add noise from a normal-distribution with mean zero and standard deviation equal to 0.1 which gives a signal-to-noise ratio of 39dB.

For the signal-to-noise ration we adopt the formula $\text{SNR}_{dB}=10\text{log}_{10}\big(\frac{P_\text{{signal}}}{P_{\text{noise}}}\big)$ where $P_\text{{signal}}$ refers to the variance observed in the normalised data and $P_{\text{noise}}$ represents the variance of the added noise. \ednote{B:Added this sentence.}
Once again consider the actual and predicted trajectories for the noisy DP dataset in Figure~\ref{fig:noisydp_success_traj}. Next we consider the densities of the predicted and actual values in Figure~\ref{fig:noisydp_success_density} and observe that they display highly similar distributions.

@@

\subsection{Additional Attractors: Clifford \& Thomas}
We follow the methodology presented in~\cite{manjunath2021universal} but opt to consider attractors different from the Lorenz system and H\'enon map in this project.

\subsubsection{Thomas' Cyclically Symmetric Strange Attractor}\label{Thomas_Attractor}
Thomas' Cyclically Symmetric Strange Attractor, a 3D attractor proposed in 1999 by Ren\'e Thomas in~\cite{ThomasAttractor}, is described by a set of three equations which is distinctly reminiscent of the Lorenz system, another deterministic three-equation model exhibiting chaotic behaviour. It has a single parameter $\beta$ and has been shown to transition to chaotic behavious when $\beta<0.208186$~\cite{Thomas_BetaParameter}.
\begin{eqnarray}\label{eqns_thomas}
  x_{n+1} = \sin(y_n) - \beta{x_n} \\
  y_{n+1} = \sin(z_n) - \beta{y_n} \\
  z_{n+1} = \sin(x_n) - \beta{z_n}
\end{eqnarray}

The behaviour of the system with a $\beta$-parameter value of 0.1056 was simulated by scaling the data from it to fit inside the interval $[-1,1]$. Two cases were considered: one noise-free and another perturbed by noise. 
Noise was generated  by adding data points taken from a normal distribution with mean 0 and standard deviation 0.05, which translates approximately to a signal-to-noise ration of 33dB. 
($\text{SNR}_{dB}$ is calculated exactly as for the DP-data above)

A sequence of 3000 observations in both the clear and noisy data-sets respectively were fed into the system with 300 entries discarded to simulate the network's loss of memory. 
The next 9000 steps were predicted and compared with true data. See Figure~\ref{fig:Thomas_trajectories} for the actual and predicted trajectories of the noise-free and perturbed systems respectively.

\begin{figure}[ht]
  \centering
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_noisy_0.1056.eps}
  \endminipage\hfill
  \minipage{0.5\linewidth}
    \centering    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_clean_0.1056.eps}
  \endminipage
 \captionof{figure}{True (red) and predicted (blue) trajectories for the Thomas attractor with parameter value $\beta=0.1056$. To the left: Noise added from normal distribution with mean zero and standard deviation 0.05. Right: No noise added to dataset.}
\label{fig:Thomas_trajectories}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.35]{Graphs/_Thomas_1.eps}\caption*{Predicted trajectories of the $x$- and $y$-coordinates for the Thomas attractor with parameter-value $\beta=0.1056$ demonstrate empirically the ability to predict the evolution of the trajectory for the next an estimated 100 timesteps into the future near-exactly. Here no noise was added.}
  \includegraphics[scale=0.5]{Graphs/_Thomas_3.eps}\caption*{Represented here the learnt (blue) and actual (red) densities of the first coordinate ($x$-coordinate) of the dataset for the Cyclically Symmetric Attractor with parameter-value $\beta=0.1056$.}
\end{figure}

@@

The behaviour of the system with a parameter values of $x_0=0$, $y_0=0$, $a = -1.7$,  $b = 1.8$, $c = -1.9$ and $d = -0.4$ was examined. 
Data was scaled to fit inside the interval $[-1,1]$. Only the noise-free case was investigated. 
Depictions of the empirical results available in Figure~\ref{fig:Clifford}.
 A sequence of 25000 observations was fed into the system with 15000 entries discarded \ednote{M: Why did you have to discard 15000?} to simulate the network's loss of memory. The next 1000 steps were predicted and compared with true data.

>>>>

In this chapter we turn to the implementation of the function $G_T$ by discussing a method to learn the function $\Gamma$. We apply our methodology to some chaotic attractors. 
Most notably, we consider the reconstruction of the chaotic attractor of the double pendulum -- an  attractor that has not previously been successfully reconstructed because of its intricate structure. In fact the previous data-driven approaches also have not been successful in building a model from the double pendulum.

@@

\section{Implementing $G_T$}
The map $G_T$ is implemented by applying the functions $\Gamma$, $g$ and the projection function $\pi_2$ in the relevant order as described by the equations~\ref{eqns_from_data}\ednote{How do I get both?}. To reach this point though, one must first learn $\Gamma$.
After first normalising the data to ensure that it has zero mean value, lies within the range $[-1,1]$, and has been zero-padded as required (see paragraph below for further discussion), we drive the system with external input through the function $g$.

From~\eqref{eqn_driving} we recall the equation of the RNN: \[g(u,x) = (1-a) x + a \tanh(Au + \alpha B x).\]
Recall that $g$ is SI-invertible when the matrix $A$ has the same dimension as $B$ and is invertible. We refer to $A$ as the \textit{input} matrix and $B$ the \textit{reservoir} matrix.

We make use of a Recurrent Neural Network (RNN) in our implementation to project the data on to a higher dimensional space. We do not train the RNN in the ways familiar to literature and opt to follow the methodology of~\cite{manjunath2021universal}. 
A few words about the RNN's: RNN's are widely used in the literature for temporal processing of data as a type of deep learning technique. 
They take the form of equations considered to be discrete-time version of continuous dynamical systems that model a neural network of neurons~\cite{jaeger2001echo}.

We implement a feedforward network (a network with no memory, unlike the RNN) to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library .

If the RNN we will be using has $N$ neurons, then $A$ and $B$ are each $N\times{N}$ matrices. Inputs $u_n$ of dimension $K < N$ are padded by zeroes (as remarked earlier) to embed the input in a space of dimension $N$. 
As we do not need the entire left-infinite history of an input, an arbitrary initialisation of the system with initial values, followed by the driving of the system with the function $g$ will mean that the generated values approach the actual solution elements in a uniform manner. This follows from the UAP that discussed in the preceding chapter (see~\ref{Dfn_UAP} and subsequent discussion).

Thirdly, we prepare the network by setting up a network with $N$ neurons  to obtain successive state values as samples in the space $Y_T$. 
The actual learning of $\Gamma$ is done using the Adam Optimiser where the means square error (MSE) loss function is optimised. 
The system is trained several times (in our case 4) with incrementally smaller learning rates on each run (0.001 and thereafter divided by 10 at each run). 
Additional parameters such as the number of network-layers, neurons per layer and training-length are presented case-by-case below.

Once $\Gamma$ is learnt, we progress to the the prediction step where $u_n$ is forecasted from $(x_{n-1},x_{n})$. Note that  due to causal embedding $(x_{n-1},x_{n})$ is in one-to-one correspondence with the left-infinite orbit $(\ldots,u_{n-1},u_{n-1})$. 
When $u_n$ is computed using $\Gamma$, it is fed to the driven system $g$ along with  $x_n$ to obtain $x_{n+1}$. Now we apply $\Gamma$ on $(x_n,x_{n+1})$ to forecast $u_{n+1}$ and the whole process repeats. Note that this corresponds to implementing the two model equations for learning from data~\eqref{eqns_from_data}.

@@

We also consider a more general problem of learning an attractor of dynamical system when only observations of an orbit are given. 
Explicitly, if $(W,T)$ is a dynamical system with dynamics generated by $w_{n+1}=Tw_n$, and if $\theta:W \to \mathbb{R}$ is an observable, then the task is to learn a dynamical system that is topologically conjugate to $(W,T)$ and predict $\theta(w_{m+1}),\theta(w_{m+2}),\ldots$ using the data $\theta(w_{0}),\theta(w_{1}),\ldots,\theta(w_{m})$.

So how do we do it? Suppose the input generated from the delay-coordinate map $\Phi_{\theta,2d}(\theta(w_{n})) := (\theta(w_{n-2d}),\ldots,\theta(w_{n-1}),\theta(w_{n}))$ is such that Takens delay embedding theorem (see Theorem~\ref{Thm_Takens}) holds.
In that case, there exists a homeomorphism  $F_\theta: \Phi_{\theta,2d}(\theta(w_{n})) \mapsto \Phi_{\theta,2d}(\theta(w_{n+1}))$. If we then feed the input values $u_n := \Phi_{\theta,2d}(\theta(w_{n}))$ to a driven system $g$ having SI-invertibility and the USP, the induced system $(Y_F,G_F)$ would be topologically conjugate to the inverse-limit space of
$(\Phi_{\theta,2d}(W), F_\theta)$ due to Theorem~\ref{Thm_CET}, and hence one could forecast the values ($u_n,u_{n+1},\ldots$).

In summary, one could forecast  the values $\theta(w_n), \theta(w_{n+1}),\ldots$ by feeding these observations as delay coordinates since the input is again an orbit from a dynamical system. 
The advantage of feeding delay-coordinates to a driven system over learning a map that describes the evolution of the delay coordinates is that the embedding is stable in the sense of global dissipativity that we have described in a previous section~\ref{subs_LearnGamma}.
I.e., there is an attractor $A$ of a map that implements $G_T$ on an expansion $Y_F^+$ providing some robustness to noise.

We now turn our attention to another practical issue related to the forecasting of chaotic dynamical systems.  
Recall that the twin equations~\ref{eqns_from_data} generate a model from data. 
For a chaotic system, a small amount of noise due either to input-noise or computational error would cause solutions from even a very accurate model with the same initial conditions to diverge. %\ednote{Grammarly}
Statistically, however,  a chaotic system will behave well in the sense that the visiting frequency (the average time spent by an orbit in any region of the phase space of the dynamical system) is identical for almost all initial conditions. 
The phrase `almost all' here is specified as a measure and refers to a result in ergodic theory (\cite{wikiErgodicTheory}) which is a topic beyond the scope of this project.
To illustrate the accuracy of the twin  equations~\ref{eqns_from_data} in modelling the data, we plot the densities of the trajectories of the actual and predicted solutions in ~\ref{fig:dp_success_density},~\ref{fig:noisydp_success_density} and ~\ref{fig:Thomas_TrajDensity}.

In our implementation, the input and reservoir  matrices $A$ and $B$ in the RNN are sparse matrices. This sparsity is not essential but seems to perform slightly empirically better than choosing the matrices arbitrarily.  We also discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opt to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of 1. The condition number of a matrix $M$ is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed.  We do not have any theoretical justifications for this.

Lastly, we ensure the matrix $B$ has unit spectral radius so that the variable $\alpha$ in the RNN equation \eqref{eqn_driving} can be used to know if the USP is satisfied. 
It is a  feature that a matrix with a spectral radius less than 1 generally has the USP~\cite{yildiz2012re}.

We remark on one more aspect that helps produce slightly better empirical results.  To learn $\Gamma: (x_{n-1},x_{n}) \mapsto u_n$ through our feedforward network, we could feed the principal components. 
Principal components are used only for a more efficient state representation to possibly reduce learning errors; it is not a lossy approximation since we use all principal components.

More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. 
$X_{1:N}=U\Sigma P^T$ here represents the singular value decomposition of $X_{1:N}$ ($U$ and $V$ are unitary matrices, and $\Sigma$ is a diagonal matrix) and we refer to \cite{wikiSVdecomp}. 
We then denote the principal component matrix by the symbol $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P$.
These principal components are used to train a feedforward neural network.

Denoting the row vectors of $Z_{1:N}$ by  $z_i^T, i=1,2,\ldots,N$ and the neural network by $NN$, the network is trained to learn an approximation of the map
$NN: (z_{n-1},z_n) \mapsto  u_n.
$
The learnt approximation of $NN$ can be used to approximate $\Gamma$ because 
\begin{equation}\label{Seq_RNN}
  NN\left( \begin{bmatrix} 
    P^Tx_{n-1} \\
    P^Tx_n
    \end{bmatrix}
    \right) = NN \circ 
    \begin{bmatrix}
    P^T & 0 \\
    0 & P^T 
    \end{bmatrix}\begin{bmatrix}
    x_{n-1}\\
    x_n
    \end{bmatrix} = u_n = \Gamma(x_{n-1},x_n).
\end{equation}

\section{Experimental Results}
In this section we showcase numerical results pertaining to three attractors that were simulated using the theory and described methodology. We first consider a simple physical system that exhibits complicated dynamics.
\subsection{Double Pendulum}

The Double Pendulum consists of two pendulums fastened to one another such that the system moves as a whole. The first (or higher pendulum) is attached to a fixed point while the second is attached to the endpoint of the first. (See Figure~\ref{fig:dp_setup} below.)
This system shows chaos in a striking manner -- consider once more Figure~\ref{fig:dp_sdic}.
The system has four variables: for each pendulum rod an associated angle and angular velocity. An ideal system is assumed where the rods are massless and pivots are frictionless.
\begin{figure}[ht]
  \includegraphics[width=0.8\linewidth]{Graphs/_dp_setup.eps}
  \centering
  \captionof{figure}{Pictorial representation of the setup of a DP where $\theta, m, l$ represent the angle, rod-length, and pendulum head mass respectively.}
\label{fig:dp_setup}
\end{figure}

The scalar equations of motion are derived using Newtonian physics and are reproduced from~\cite{DPFormulas} in Equations~\ref{eqns_dp}. 
Each pendulum component has an associated mass of the head, $m$, and length of the rod, $L_i$. Acceleration due to gravity is denoted by the variable $g$ and initial velocities by $\dot{\theta}$
\begin{eqnarray}\label{eqns_dp}
  \ddot{\theta_{1}}  = \frac{-g(2m_1+m_2)\sin{\theta_1} - m_2g\sin(\theta_1-2\theta_2) - 2\sin(\theta_1-\theta_2)m_2({\dot{\theta_{2}}}^{2}L_2 + {\theta_{1}'}L_1\cos(\theta_1-\theta_2))} {L_1(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
  \\
  \ddot{\theta_{2}} = \frac{2\sin(\theta_1-\theta_2)(\theta_{1}'^{2}L_1(m_1+m_2) + g(m_1+m_2)\cos\theta_1 + \dot{\theta_{2}}^{2}L_2m_2\cos(\theta_1-\theta_2))}{L_2(2m_1 + m_2 -m_2\cos(2\theta_1 - 2\theta_2))}
\end{eqnarray}

% The second-order system is converted to a first-order ODE system and solved using the Runge-Kutta numerical method.  
Here we chose the values $m_1=2.0$, $m_2=2.0$, $L_1=1.5$ and $m_2=1.5$ with initial velocities $v_1=1.5, v_2=-2$ for the respective pendulum heads.
With a change of variables the second-order system can be written as a system of first order equations with four variables, where each pendulum has an associated $x$ and $y$ coordinates of the mass (the lengthy equations are not shown here). 
The ODE system is then solved using the fourth-order Runge-Kutta numerical method.

@@

We need the assumption of an undamped system so that the map $T$ resulting from the discretization of the system  is surjective.  If we assume, damping surjectivity is lost on any phase space with more than one point.  
An example of this would be the case of a DP exhibiting energy-loss. In such a case, the attractor is a single point in the phase space representing the instant when the double pendulum comes to rest. 
If we use data outside the attractor in this scenario, learning is not reliable. To see this, consider the plots~\ref{fig:damped_pendulum} and~\ref{fig:dp_notsurjective}.

@@

The system was driven for 15500 steps, of which the first 500 were immediately discarded to simulate the network `forgetting'. 
Input and reservoir matrices of the driven system $g$ were implemented as sparse matrices filled to 20\%. Parameter-values of $\alpha=0.99$ and $a=0.5$ were used.
The remaining 15000 datapoints were used to train a network with 128 neurons. 
The network was initialised to contain 16 hidden layers each with layer dimension 64 and activated with the $\tanh$ activation function.
Training was accomplished using 256 epochs with batch sizes 128. See Fig.~\ref{fig:dp_success_traj} for results pertaining to the trajectory taken component-wise from the DP's and then Fig.~\ref{fig:dp_success_density} to witness associated densities of predicted versus actual values.

@@

We now consider an experiment where noise is added to the system and find it to be incredibly robust. Employing the same parameters as in the preceding experiment, we include noise from a normal-distribution with zero mean value and standard deviation equal to 0.1 which equates to a signal-to-noise ratio of 39dB.

For the signal-to-noise ration we adopt the formula $\text{SNR}_{dB}=10\text{log}_{10}\big(\frac{P_\text{{signal}}}{P_{\text{noise}}}\big)$ where $P_\text{{signal}}$ refers to the variance observed in the normalised data and $P_{\text{noise}}$ represents the variance of the added noise.
Once again we display the actual and predicted trajectories for the noisy DP data in Figure~\ref{fig:noisydp_success_traj}. 
Next we consider the densities of the predicted and actual values in Figure~\ref{fig:noisydp_success_density} and observe that they display highly similar distributions.

@@

\subsection{Additional Attractors: Clifford \& Thomas}
We follow the methodology presented in~\cite{manjunath2021universal} but opt to consider attractors different from the Lorenz system and H\'enon map for this project.

\subsubsection{Thomas' Cyclically Symmetric Strange Attractor}\label{Thomas_Attractor}
Thomas' Cyclically Symmetric Strange Attractor, a 3D attractor proposed in 1999 by Ren\'e Thomas in~\cite{ThomasAttractor}, is described by a set of three equations which is distinctly reminiscent of another deterministic three-equation model exhibiting chaotic behaviour proposed by Lorenz. 
It has a single parameter $\beta$ and has been shown to transition to chaotic behavious when $\beta<0.208186$~\cite{Thomas_BetaParameter}.
\begin{eqnarray}\label{eqns_thomas}
  x_{n+1} = \sin(y_n) - \beta{x_n} \\
  y_{n+1} = \sin(z_n) - \beta{y_n} \\
  z_{n+1} = \sin(x_n) - \beta{z_n}.
\end{eqnarray}

The behaviour of the system with a $\beta$-parameter value of 0.1056 was simulated by scaling the data generated from the equations to fall inside the interval $[-1,1]$. Two cases were considered: one noise-free and another perturbed by noise. 
Noise was generated  by adding data points taken from a normal distribution with a mean of zero and standard deviation 0.05, which translates approximately to a signal-to-noise ration of 33dB. 
$\text{SNR}_{dB}$ is calculated exactly as for the DP-data.

A sequence of 3000 observations in both the clear and noisy data-sets respectively was fed into the system with 300 entries discarded to simulate the network's loss of memory. 
The next 9000 steps were predicted and compared with true data. See Figure~\ref{fig:Thomas_trajectories} for the actual and predicted trajectories of the noise-free and perturbed systems respectively.

\begin{figure}[ht]
  \centering
  \minipage{0.5\linewidth}
    \centering
    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_noisy_0.1056.eps}
  \endminipage\hfill
  \minipage{0.5\linewidth}
    \centering    \includegraphics[width=0.8\linewidth]{Graphs/_thomas_clean_0.1056.eps}
  \endminipage
 \captionof{figure}{True (red) and predicted (blue) trajectories for the Thomas attractor with parameter value $\beta=0.1056$.Left: Noise added from normal distribution with mean zero and standard deviation 0.05. Right: No noise added to dataset.}
\label{fig:Thomas_trajectories}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.35]{Graphs/_Thomas_1.eps}\caption*{Predicted trajectories of the $x$- and $y$-coordinates for the Thomas attractor with parameter-value $\beta=0.1056$ demonstrate empirically the ability to predict the evolution of the trajectory for the next an estimated 100 timesteps into the future near-exactly. Here no noise was added.}
  \includegraphics[scale=0.5]{Graphs/_Thomas_3.eps}\caption*{Represented here the learnt (blue) and actual (red) densities of the first coordinate ($x$-coordinate) of the dataset for the Cyclically Symmetric Attractor with parameter-value $\beta=0.1056$.}
\end{figure} \label{fig:Thomas_TrajDensity}

@@
